Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2024-01-04 paper/paper-new.Rmd
update

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      dev = "pdf",
                      fig.align = "center",
                      size = "scriptsize",
                      message = FALSE,
                      warning = FALSE)
```

@@

```{r include = FALSE}
# figures
source(here("scripts", "figures.R"), local = knitr::knit_global())
rfigs <- readRDS(here("objects", "r-figs.rds"))

# tables
source(here("scripts", "figures.R"), local = knitr::knit_global())
rtabs <- readRDS(here("objects", "r-tabs.rds"))
```

@@

Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions. We can name this general class of models as *ordinal regression*. The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures [@Tutz2022-dg]. @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models. We can identify three main classes: *cumulative models* [CM; @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe] and *adjacent category models*. The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.

@@

In the current paper we put the focus on the CM for several reasons. The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation. The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item. Furthermore, CM are also used to model data under a signal detection theory framework [e.g., @DeCarlo2010-lj]. The Figure \@ref(fig:fig-ordinal-models) depict the overall structure of the *cumulative* model.

```{r fig-ordinal-models, fig.cap=filor::get_caption(), out.width="100%"}
knitr::include_graphics("img/fig-ordinal-models.png")
```

@@

```{=tex}
\begin{equation} 
P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
(\#eq:prob-cum-model1)
\end{equation}
```

The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g()$. To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:eq:prob-cum-model2).

```{=tex}
\begin{equation} 
P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta), \;\;k = 1, \dots, k - 1
(\#eq:prob-cum-model2)
\end{equation}
```
There are always two special cases when computing the probability of a single outcome $Y$ that is when $Y = 1$ and $Y = k$. In the first case the cumulative probability is calculated from $-\infty$ that is the same as temporary assuming an $\alpha_0 = -\infty$. In the second case ($Y = 1$) the probability is calculated as $P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)$ that is the same as assuming a temporary threshold $\alpha_k = +\infty$. The Figure \@ref(fig:fig-explain-cumulative) shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.

@@

```{=tex}
\begin{equation} 
Y^\star_i = \eta + \epsilon_i
(\#eq:latent-model)
\end{equation}
```
The crucial part is $\epsilon_i$ that is the random component of the model coming from a certain probability distribution. For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution. Following the notation by @Tutz2022-dg, the observed ordinal value $Y_i = k$ comes from $Y^\star_i$ belonging to the interval defined by the thresholds $Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}$ where $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.

@@

```{r fig-logit-vs-probit, fig.cap=filor::get_caption()}
dplot <- ggplot(data = NULL) +
  stat_function(fun = dnorm, aes(color = "Normal"), lwd = 1) +
  stat_function(fun = dlogis, aes(color = "Logistic"), lwd = 1) +
  xlim(c(-7, 7)) +
  theme(legend.position = "bottom")

pplot <- ggplot(data = NULL) +
  stat_function(fun = pnorm, aes(color = "Normal"), lwd = 1) +
  stat_function(fun = plogis, aes(color = "Logistic"), lwd = 1) +
  xlim(c(-7, 7))

plot_grid_common_legend(
  dplot, 
  pplot + theme(axis.title.y = element_blank()), nrow = 2
)
```

@@

```{r dataset-example1}

b1 <- log(3) # log odds ratio
k <- 4
n <- 100
x <- rep(c("a", "b"), each = n/2)
dat <- data.frame(x = x)
probs <- rep(1/k, k) # for the group "a", uniform probabilities
dat <- sim_ord_latent(~x, By = b1, prob = probs, data = dat, link = "logit")

funs <- list(mean = mean, median = median, sd = sd)
ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
ff <- t(table(dat$y, dat$x))

ss <- data.frame(cbind(ss, ff))
names(ss) <- c("mean", "median", "sd", paste0("Y", 1:4))

ss$tot <- apply(ss[, 4:ncol(ss)], 1, sum)
pp <- ss[, 4:7]/50
ss <- round(ss, 2)
ss$Y1 <- sprintf("**%s** (*p = %.2f*)", ss$Y1, pp$Y1)
ss$Y2 <- sprintf("**%s** (*p = %.2f*)", ss$Y2, pp$Y2)
ss$Y3 <- sprintf("**%s** (*p = %.2f*)", ss$Y3, pp$Y3)
ss$Y4 <- sprintf("**%s** (*p = %.2f*)", ss$Y4, pp$Y4)

ss$group <- rownames(ss)
ss <- select(ss, group, everything(), -tot)
ss
```

```{r tab-dataset-example1}
#| eval: false

flextable(ss) |> 
  autofit() |> 
  set_caption(caption = filor::get_caption()) |> 
  align(part = "all", align = "center") |> 
  ftExtra::colformat_md()
```

```{r model-example-1}
fit <- clm(y ~ x, data = dat, link = "logit")
summary(fit)
```

The two main sections of the model summary are the *Coefficients* section reporting the regression coefficients $\beta$ and the *Threshold* section reporting the $\alpha$ estimation. Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect. As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht][^3]

@@

```{r}
#| echo: true
#| collapse: true

@@

```{r}
#| echo: true
#| collapse: true
# the dummy_ord() is a custom function creating k - 1 dummy variable for the cumulative probabilities

@@

```{r}
#| echo: true
#| collapse: true

@@

# extracting the predicted probabilities for the two groups
pr <- predict(fit, data.frame(x = unique(x)))$fit

@@

The proportional odds assumption is relevant only for the *logit* model because of the link function that put $\beta$ as log odds. However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model. Equation \@ref(eq:probit-prop-odds) depict the parallel slopes assumption for a probit model. Basically the difference in $z$ for a unit increase in $x$ (i.e., the slope) is the same regardless the $Y$ level. This can be directly demonstrated in the same way as for the logit model. We use the same dataset presented in Table \@ref(tab:tab-prop-ord-ex) but fitting a probit model. The difference between the $z$ scores comparing the two groups calculated on the cumulative probability is the same regardless the $k - 1$ level.

$$
z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
$$

```{r}
#| echo: true
#| collapse: true

@@

The main difference regards the parameters interpretation. In the logit model the $\beta$ are the log odds ratio. For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in $x$. For the probit model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$. This is very convenient especially for categorical variables because parameters can be intepreted as a Cohen's $d$ like measure. Thinking about the latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$. The intepretation in terms of shifting the latent mean holds also for the logistic model. However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$[^4]. The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\frac{\pi}{\sqrt{3}}$ standard deviations.

[^4]: Actually the variance of the logistic distribution is $\frac{s^2\pi^2}{3}$ and the standard deviation $\frac{s\pi}{\sqrt{3}}$ where $s$ is the scale of the distribution. For the standard logistic distribution $s = 1$ (as for the standard normal distribution). For the standard normal distribution $s^2 = \sigma^2 = 1$

Another advantage of the probit model is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into the model coefficients. In practical terms, the thresholds $\alpha$ are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].

qualcosa in più qui

@@

There are mainly two ways to simulate data. The first method concerns simulating starting from the latent formulation of the ordinal model. Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.

@@

The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution[^5]. This is similar to the general method to simulate data for a generalized linear model where the linear predictors $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component. The following code box simulate a $n$ binary trials with a continuous predictor $x$.

[^5]: In R we are using the `sample()` function that simulate values from a *categorical* distribution. The *categorical* distribution is considered a special case of the multinomial distribution

```{r}
xl <- seq(0, 1, 0.1)
x <- rep(xl, 10)
n <- length(x)
b0 <- qlogis(0.01)
b1 <- 8
lp <- b0 + b1 * x # linear predictor
y <- rbinom(n, 1, plogis(lp)) # sampling from a binomial distribution using the vector of probabilities
p <- tapply(y, x, mean)
plot(unique(x), p, pch = 19)
curve(plogis(x, -(b0/b1), 1/b1), add = TRUE, lwd = 2, col = "firebrick")
```

Essentially the `lp` ($\eta_i$) vector contains true probability of success ($p_i|x_i = g^{-1}(\eta_i)$) for the $x_i$ level of the predictor. Then to introduce the random part of the model, we can use $p_i$ to generate binary values from a Binomial (in this specific case a Bernoulli) distribution. This create `y` that is the vector of 0-1 values generated using the true probabilities of success. The supplementary materials contains a more extensive example for simulating data for generalized linear models.

We can apply the same idea to an ordinal outcome but we need $k - 1$ equations (compared to the previous example) where $k$ is the number of ordinal levels. Let's simulate a similar design with a continuous $x$ predictor and $k = 4$ options. We fix the baseline probabilities where $x = 0$ as uniform thus $p(y_1) = p(y_2) = ... p(y_k) = 1/k$. The proposed workflow for the simulation is:

1.  Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
2.  Define the regression coefficients $\beta$s.
3.  Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
4.  Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
5.  Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
6.  Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities. For example, the probability of $P(y = 2) = P(y \leq 2) - P(y \leq 1)$. This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities because the probability of the first $Y = 1$ level is the cumulative probability from $- \infty$ to the first threshold and the probability of $Y = k$ is obtained removing $P(Y \leq k - 1)$ from one.
7.  Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.

```{r}
k <- 4  # number of options (1)
n <- 1e5 # number of observations (1)
b1 <- 0.5 # beta1, the shift in the latent distribution (2)
probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
dat <- data.frame(x = rnorm(n)) # create dataframe
lp <- lapply(alphas, function(a) a - b1 * dat$x) # k - 1 linear predictors (4)
names(lp) <- sprintf("g[p(y <= %s)]", 1:(k - 1)) # giving appropriate names
```

The linear predictors ($\eta$) can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.

```{r}
#| echo: false
lpt <- lp
names(lpt) <- latex2exp::TeX(sprintf("$g[p(y \\leq %s)]$", 1:(k - 1)))
datl <- cbind(dat, lpt)
datl <- tidyr::pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")

ggplot(datl, aes(x = x, y = lp, color = cumy)) +
  facet_wrap(~cumy, labeller = label_parsed) +
  geom_hline(yintercept = alphas, lty = "dashed") +
  geom_line(show.legend = FALSE) +
  theme_minimal(20) +
  ylab("Linear Predictor")
```

Now we can apply the inverse of the link function $g^{-1} = \Phi^{-1}$ to calculate the corresponding cumulative probabilities.

```{r}
cump <- lapply(lp, pnorm) # inverse of the link function (5)
cump <- data.frame(cump)
p <- apply(cbind(0, cump, 1), 1, diff, simplify = FALSE) # probability of each k outcome (6)
p <- do.call(rbind, p)
```

```{r}
p <- data.frame(p)
names(p) <- sprintf("p_y%s", 1:k)
filor::trim_df(data.frame(p))
```

We can plot the expected effect of $\beta_1$ on the $k$ probabilities using the `num_latent_plot()` function (see Figure \@ref(fig:sim-with-probs-lat-plot)). Finally we sample using the `sample()` function using the calculated probabilities.

```{r sim-with-probs-lat-plot, fig.cap="sim-with-probs-lat-plot"}
num_latent_plot(x = dat$x, b1 = b1, probs = probs0, link = "probit")
```

```{r}
dat$y <- apply(p, 1, function(ps) sample(1:k, 1, prob = ps))
```

To check the simulation result we can increase the number of observations, fit the model using the `ordinal::clm()` function and assess the recovery of simulated parameters.

```{r}
dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
fit <- clm(y ~ x, data = dat, link = "probit")
```

```{r tab-sim-probs}
truth <- c(alphas, b1)
names(truth) <- c(names(fit$alpha), names(fit$beta))
#clm_table(fit, truth = truth)
```

@@

```{r}
k <- 4  # number of options (1)
n <- 1e5 # number of observations (1)
b1 <- 0.5 # beta1, the shift in the latent distribution (2)
probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
dat <- data.frame(x = rnorm(n)) # create dataframe
dat$ystar <- b1 * dat$x + rnorm(n, 0, 1) # linear predictor (4)
dat$y <- findInterval(dat$ystar, alphas) + 1 # cut the latent distribution (5). The + 1 because the first category is 0 by default.
```

The Figure (put crossref) depict the simulated $Y^{*}$ and the corresponding ordinal value. As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.

```{r}
alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))

dat |> 
  slice_sample(n = 1000) |> 
  mutate(y = findInterval(ystar, alphas) + 1) |> 
  ggplot(aes(x = x, y = ystar)) +
  geom_point(alpha = 0.6, aes(color = factor(y)), size = 3) +
  theme_minimal(15) +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  ylab(latex2exp::TeX("$Y^{*}$")) +
  geom_hline(yintercept = alphas) +
  annotate("label", x = min(dat$x), y = alphas, label = alpha_n)
```

```{r}
dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
fit <- clm(y ~ x, data = dat, link = "probit")
```

```{r tab-sim-latent}
truth <- c(alphas, b1)
names(truth) <- c(names(fit$alpha), names(fit$beta))
#clm_table(fit, truth)
```

The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function.

```{r}
#| results: asis
filor::print_fun(funs$sim_ord_latent)
```

@@

A common research design could be a 2x2 factorial design. In this example we have 2 main effects and the interaction. Regardless the link function, we can focus the model equation on the linear predictor. Equation depicts the linear predictor for a 2x2 interaction. With a 2x2 design, a good strategy is to use sum to zero contrasts for the two factors. By default R use dummy coding but here we set the contrasts for $X_1$ and $X_2$ as $0.5$ and $-0.5$. In this way $\beta_1$ will be the main effect of $X_1$, $\beta_2$ the main effect of $X_2$ and $\beta_3$ the interaction (thus the difference of differences).

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      dev = "pdf",
                      fig.align = "center",
                      size = "scriptsize",
                      collapse = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      comment = "#>")
```

@@

```{r scripts, include = FALSE}
temp <- new.env()
# figures
# source(here("scripts", "figures.R"), local = temp)
rfigs <- readRDS(here("objects", "r-figs.rds"))

# tables
# source(here("scripts", "tables.R"), local = temp)
rtabs <- readRDS(here("objects", "r-tabs.rds"))
```

```{r paper-functions}
head_tail <- function(x, n = 5){
  rbind(
    head(x, n = n),
    tail(x, n = n)
  )
}
```

@@

Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions. We can name this general class of models as *ordinal regression*. The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures [@Tutz2022-dg]. @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models. We can identify three main classes: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe] and *adjacent category models*. The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.

@@

In the current paper we put the focus on the CM for several reasons. The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation. The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item. Furthermore, CM are also used to model data under a signal detection theory framework [e.g., @DeCarlo2010-lj]. The Figure \@ref(fig:fig-explain-cumulative) depict the overall structure of the *cumulative* model.

@@

\begin{equation} 
P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
(\#eq:prob-cum-model1)
\end{equation}

The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g()$. To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:prob-cum-model2).

\begin{equation} 
P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta), \;\;k = 1, \dots, k - 1
(\#eq:prob-cum-model2)
\end{equation}

There are always two special cases when computing the probability of a single outcome $Y$ that is when $Y = 1$ and $Y = k$. In the first case the cumulative probability is calculated from $-\infty$ that is the same as temporary assuming an $\alpha_0 = -\infty$. In the second case ($Y = 1$) the probability is calculated as $P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)$ that is the same as assuming a temporary threshold $\alpha_k = +\infty$. The Figure \@ref(fig:fig-explain-cumulative) shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.

@@

\begin{equation} 
Y^\star_i = \eta + \epsilon_i
(\#eq:latent-model)
\end{equation}

The crucial part is $\epsilon_i$ that is the random component of the model coming from a certain probability distribution. For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution. Following the notation by @Tutz2022-dg, the observed ordinal value $Y_i = k$ comes from $Y^\star_i$ belonging to the interval defined by the thresholds $Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}$ where $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.

@@

```{r fig-logit-vs-probit, fig.cap=filor::get_caption()}
dplot <- ggplot(data = NULL) +
  stat_function(fun = dnorm, aes(color = "Normal"), lwd = 1) +
  stat_function(fun = dlogis, aes(color = "Logistic"), lwd = 1) +
  xlim(c(-7, 7)) +
  theme(legend.position = "bottom") +
  viridis::scale_color_viridis(discrete = TRUE)

pplot <- ggplot(data = NULL) +
  stat_function(fun = pnorm, aes(color = "Normal"), lwd = 1) +
  stat_function(fun = plogis, aes(color = "Logistic"), lwd = 1) +
  xlim(c(-7, 7)) +
  viridis::scale_color_viridis(discrete = TRUE)

legend <- cowplot::get_legend(dplot + theme(legend.position = "bottom", legend.title = element_blank()))
grid <- cowplot::plot_grid(dplot + rm_legend(), pplot + rm_legend())
cowplot::plot_grid(grid, legend, nrow = 2, rel_heights = c(0.9, 0.1))
```

@@

```{r tab-dataset-example}
dat <- rtabs$data_example_table$dat
map(rtabs$data_example_table$table, ~gsub("\\**", "", .x)) |> 
  bind_rows() |> 
  kable(row.names = FALSE,
        booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE) |> 
  kable_styling(full_width = FALSE)
```

```{r model-example-1}
#| echo: true
fit <- clm(y ~ x, data = dat, link = "logit")
summary(fit)
```

The two main sections of the model summary are the *Coefficients* section reporting the regression coefficients $\beta$ and the *Threshold* section reporting the $\alpha$ estimation. Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect. As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [see @Schad2020-ht][^3]

@@

```{r odds-example}
#| echo: true
#| collapse: true

@@

```{r cumulative-odds-example}
#| echo: true
#| collapse: true
# the dummy_ord() is a custom function creating k - 1 dummy variable for the cumulative probabilities

@@

```{r poa-logit-example}
#| echo: true
#| collapse: true

@@

# extracting the predicted probabilities for the two groups
pr <- predict(fit, data.frame(x = unique(dat$x)))$fit

@@

The proportional odds assumption is relevant only for the *logit* model because of the link function that put $\beta$ as log odds. However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model. Equation \@ref(eq:probit-prop-odds) depict the parallel slopes assumption for a probit model. Basically the difference in $z$ for a unit increase in $x$ (i.e., the slope) is the same regardless the $Y$ level. This can be directly demonstrated in the same way as for the logit model. We use the same dataset of the previous simulation. The difference between the $z$ scores comparing the two groups calculated on the cumulative probability is the same regardless the $k - 1$ level.

\begin{equation}
z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
(\#eq:probit-prop-odds)
\end{equation}

```{r poa-probit-example}
#| echo: true
#| collapse: true

@@

The main difference regards the parameters interpretation. In the logit model the $\beta$ are the log odds ratio. For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in $x$.

For the *probit* model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$. This is very convenient especially for categorical variables because parameters can be interpreted as a Cohen's $d$ like measure. Thinking about the latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$. More formally the shift in the latent distribution is $\frac{\beta}{\sigma}$ (for the *probit* model $\sigma = 1$). The interpretation in terms of shifting the latent mean holds also for the logistic model. However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$[^4]. The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\beta/(\frac{\pi}{\sqrt{3}})$ standard deviations.

[^4]: Actually the variance of the logistic distribution is $\frac{s^2\pi^2}{3}$ and the standard deviation $\frac{s\pi}{\sqrt{3}}$ where $s$ is the scale of the distribution. For the standard logistic distribution $s = 1$ (as for the standard normal distribution).

@@

In this tutorial we present two method for simulating ordinal data. The first method concerns calculating the probabilities of each $k$ of $Y$ as a function of predictors and then sampling from a multinomial distribution. The second method simulate the latent variable $Y^\star$ as a function of predictors.

@@

For the first method we need to calculate $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution^[In R we are using the `sample()` function that simulate values from a *categorical* distribution. The *categorical* distribution is considered a special case of the multinomial distribution]. This is similar to the general method to simulate data for a generalized linear model where the linear predictor $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.

As a simple example, let's simulate two groups or condition with $n = 100$ participants responding to an item with $k = 4$ ordered options. We simulate that the second group has more probability of responding higher categories of $Y$. As explained in the previous sections, we can summarise the effect size of a CM (assuming POA) using a single $\beta = \log(OR)$. We can assume that the odds ratio is $OR = 2$. For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = ... P(Y = k|x_0) = 1/k$. The following code summarise the first steps of the simulation. Basically we define the simulation parameters, we calculate the $k - 1$ thresholds $\alpha$ and we apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2). In this way we calculated the true probability of each $Y$ level for the two groups.

An important step is converting from $\alpha$ to probability and the opposite. This steps are implemented in the `alpha_to_prob()` and `prob_to_alpha()` functions. Basically given the input and the link function, these functions return the thresholds associated with $k$ probabilities or the $k - 1$ thresholds.

```{r alpha-prob-example}
#| echo: true
#| collapse: true
k <- 5
(p <- rep(1/k, k)) # uniform
names(p) <- paste0("y", 1:k)

(alpha <- prob_to_alpha(p, link = "logit")) # or prob_to_alpha(p, "probit")
alpha_to_prob(alpha, link = "logit")
```

```{r simulating-from-multinomial1}
#| echo: true
#| collapse: true

## SIMULATION PARAMETERS

N <- 100 # sample size
or <- 2 # odds ratio
k <- 4 # number of ordinal alternatives
probs0 <- rep(1/k, k) # probabilities for the first group
alpha <- prob_to_alpha(probs0, link = "logit")
dat <- data.frame(x = rep(c(0, 1), each = N/2))

## LINEAR PREDICTOR

# calculate linear predictor using equation TODO put equation
# obtaining k - 1 equations

lp <- lapply(alpha, function(a) a - log(or) * dat$x)
names(lp) <- sprintf("lp_leq%s", 1:(k - 1)) # giving appropriate names
lp <- data.frame(lp)
head(lp)

## CUMULATIVE PROBABILITIES

# apply the inverse of the link function (invlogit) to calculate cumulative probabilities
cump <- lapply(lp, plogis)
cump <- data.frame(cump)
names(cump) <- sprintf("cump_leq%s", 1:(k - 1)) # giving appropriate names
head(cump)

## PROBABILITIES OF Y

# for each row, we can calculate P(Y = k) using equation TODO put equation
# P(Y = 1) = P(Y <= 1)
# P(Y = 2) = P(Y <= 2) - P(Y <= 1)
# P(Y = 3) = P(Y <= 3) - P(Y <= 2)
# P(Y = 4) = 1 - P(Y <= 3)

# adding a columns of 0 and 1, then diff() for adjacent differences
cump <- cbind(0, cump, 1)
p <- apply(cump, 1, diff, simplify = FALSE)

p <- data.frame(do.call(rbind, p)) # collapse list of rows into a dataframe
names(p) <- sprintf("p%s", 1:k) # giving appropriate names
head(p)

# probabilities for id = 1 (x = 0) and id = 51 (x = 1)
p[c(1, 51), ]

## CUMULATIVE ODDS RATIO

# calculate the (cumulative) odds ratio

x0 <- cump[1, 2:k]
x1 <- cump[51, 2:k]

# this is the same as the or (the b1) and we are assuming POA
odds(x0) / odds(x1)
```

```{r simulating-from-multinomial2}
#| echo: true
#| collapse: true

## SAMPLING FROM THE MULTINOMIAL DISTRIBUTION

# example of a random Y outcome based on the probabilities
sample(x = 1:k, size = 1, prob = p[1, ])

# we can apply it to the full dataset, this step lead to different results
# each time we run the code because we are sampling from a distribution

dat$y <- apply(p, 1, function(ps) sample(1:k, size = 1, prob = ps))
head(dat)

# let's compute the observed probabilities, to be compared to the true
# probabilities

# observed
(op <- prop.table(table(dat$x, dat$y), margin = 1))

# true (just selecting a row from x = 0 and x = 1)
p[c(1, 51), ]

# similarly we can compute the observed cumulative odds ratios

(cum_op <- apply(op, 1, cumsum))
odds(cum_op[, 1]) / odds(cum_op[, 2])

# the odds ratios are not the same as the parameter. as we increase N
# the parameter will converge to the true value
```

```{r simulating-from-multinomial3}
#| echo: true
#| collapse: true
dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
fit <- clm(y ~ x, data = dat, link = "probit")
summary(fit)
```

We can generalize the previous workflow as:

1.  Define simulation parameters, baseline probabilities, etc. as in the previous workflow
4.  Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
5.  Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
6.  Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
7.  Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.
8.  Fit the appropriate model using `ordinal::clm()`

A general suggestion is to increase the sample size $n$ to a very large value reducing the random variability to check if simulated parameters are recovered.

@@

```{r simulating-from-latent1}
#| echo: true
#| collapse: true

## SIMULATION PARAMETERS

N <- 1e3 # sample size
or <- 4 # odds ratio, higher here just for a more clear plot
k <- 4 # number of ordinal alternatives
probs0 <- rep(1/k, k) # probabilities for the first group
alpha <- prob_to_alpha(probs0, link = "logit")
dat <- data.frame(x = rep(c(0, 1), each = N/2))

## LINEAR PREDICTOR

# calculate the linear predictor using equation TODO put equation
dat$lp <- log(or) * dat$x

# add the random part by sampling errors from a standard logistic (or normal) distribution
dat$ystar <- dat$lp + rlogis(N, location = 0, scale = 1)

# cut the latent distribution. The + 1 because the first category is 0 by default.
dat$y <- findInterval(dat$ystar, alpha) + 1

head_tail(dat, n = 3)
```

The Figure \@ref(fig:fig-sim-from-latent) depict the simulated $Y^{*}$ and the corresponding ordinal value. As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.

```{r fig-sim-from-latent}
alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))

dat |> 
  ggplot(aes(x = factor(x), y = ystar)) +
  geom_point(aes(color = factor(y)),
             position = position_jitter(width = 0.1),
             alpha = 0.7) +
  geom_hline(yintercept = alpha) +
  annotate("label", x = 0.5, y = alpha, label = alpha_n) +
  ylab(latex2exp::TeX("$Y^{*}$")) +
  xlab("Group (x)") +
  mtheme() +
  viridis::scale_color_viridis(discrete = TRUE, direction = -1) +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

```{r simulating-from-latent2}
#| echo: true
#| collapse: true
dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
fit <- clm(y ~ x, data = dat, link = "logit")
summary(fit)
```

The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function. Details about the function can be found on the Github repository LINK.

```{r}
#| echo: true
#| collapse: true
N <- 100 # sample size
or <- 4 # odds ratio, higher here just for a more clear plot
k <- 4 # number of ordinal alternatives
probs0 <- rep(1/k, k) # probabilities for the first group
alpha <- prob_to_alpha(probs0, link = "logit")
dat <- data.frame(x = rep(c(0, 1), each = N/2))

# same as the previous simulation
dat <- sim_ord_latent(~x, By = log(or), prob = probs0, data = dat, link = "logit")

head_tail(dat, n = 3)
```

@@

A common research design could be a 2x2 factorial design. In this example we have 2 main effects and the interaction. Regardless the link function, we can focus the model equation on the linear predictor. Equation depicts the linear predictor for a 2x2 interaction. With a 2x2 design, a good strategy is to use sum to zero contrasts for the two factors. By default R use dummy coding but here we set the contrasts for $X_1$ and $X_2$ as $0.5$ and $-0.5$. In this way $\beta_1$ will be the main effect of $X_1$, $\beta_2$ the main effect of $X_2$ and $\beta_3$ the interaction (thus the difference of differences). Equation \@ref(eq:interaction-equation) depict the model formula.
